Space and complexities of territorial systems \\\bigskip \textit{Actes des Journ{\'e}es de Rochebrune 2019}

### 2019-01-26 Docs/Rochebrune2019/Paper/paper_en.tex
traduction

\title{Espace et complexités des systèmes territoriaux
\\\bigskip
\textit{Actes des Journ{\'e}es de Rochebrune 2019}
}
\author{\noun{Juste Raimbault}$^{1,2,3}$
\bigskip\\
$^1$ UPS CNRS 3611 ISC-PIF\\
$^2$ CASA, UCL\\
$^3$ UMR CNRS 8504 G{\'e}ographie-cit{\'e}s
}
\date{}

@@

\begin{abstract}
	Le caractère spatialisé des systèmes territoriaux joue un rôle déterminant dans l'émergence de leurs complexités. Cette contribution vise à illustrer dans quelle mesure différents types de complexités peuvent se manifester dans des modèles de tels systèmes. Nous développons d'un point de vue théorique des arguments illustrant la complexité ontologique, au sens de la multitude et multidimensionalité de représentations possibles, puis la complexité au sens de l'émergence, c'est-à-dire la nécessité de l'existence de plusieurs niveaux autonomes. Nous proposons ensuite des expériences numériques pour explorer des propriétés de la complexité (complexité dynamique et co-évolution) au sein de deux modèles simples de morphogenèse urbaine. Nous suggérons finalement d'autres dimensions de la complexité qui pourraient être typiques des systèmes territoriaux.\medskip\\
	\noindent\textbf{Mots-clés : }\textit{Complexités; Systèmes territoriaux; Morphogenèse urbaine; Co-évolution}
\end{abstract}

@@

Les systèmes territoriaux, qui peuvent être compris comme des structures socio-spatiales auto-organisées \citep{pumain1997pour}, sont une illustration typique de systèmes complexes étudiés sous de nombreux angles incluant plus ou moins l'aspect spatial de ces systèmes. Notre appréhension des systèmes territoriaux se place plus particulièrement dans la lignée de la théorie évolutive des villes~\citep{pumain2018evolutionary} qui comprend les systèmes urbains comme des systèmes multi-niveaux, dans lesquels la co-évolution des multiples composants et agents détermine la dynamique de ceux-ci~\citep{raimbault2018caracterisation}. La complexité de ces systèmes est ainsi étroitement liée à leur caractère spatial, puisque leurs dynamiques sont portées par les distributions spatiales de leur entités et sous-systèmes, et que les interactions entre agents conduisant au comportement émergent sont inscrites dans l'espace.

Le concept de complexité correspond à de diverses approches et définitions, qui dépendent des domaines où elles sont introduites. Par exemple, \cite{chu2008criteria} passe en revue les approches conceptuelles et opérationnelles liées au champ de la vie artificielle et montre qu'il n'existe pas de concept unifié. \cite{deffuant2015visions} développe aussi différentes visions de la complexité, allant d'une complexité proche d'une complexité computationnelle à une complexité irréductible propre aux systèmes sociaux. Un précédent travail par \cite{raimbault2018relating} s'était proposé de suggérer des ponts épistémologiques entre différentes approches et définitions de la complexité, et plus particulièrement les liens entre complexité au sens d'émergence, complexité computationnelle et complexité informationnelle. Comme le souligne \cite{batty2018defining}, les dimensions de la complexité sont potentiellement infinies et il ne peut exister d'approche universellement valide pour l'ensemble des problèmes et systèmes. Il est alors important de combiner différentes approches de la complexité pour comprendre son rôle au sein des systèmes territoriaux.

Cette contribution vise à illustrer une approche des systèmes territoriaux par la géographie urbaine, et dans quelle mesure leur complexité est intimement liée à leur caractère spatial. Nous tâchons ici d'illustrer les liens entre complexité et espace selon différentes vues de celle-ci, à la fois à travers des considérations théoriques mais aussi par l'exploration de modèles de simulation de systèmes territoriaux. La suite de cet article est organisée de la façon suivante : dans deux premières sections nous développons une approche conceptuelle de la complexité ontologique et de l'émergence au sein des systèmes territoriaux. Nous présentons ensuite des résultats de simulation d'un modèle de morphogenèse urbaine exhibant des propriétés typiques de la complexité dynamique. Une autre expérience permet ensuite de montrer le lien entre complexité au sens d'une co-évolution et la structure spatiale du système. Nous discutons finalement les implications des ces résultats et des développements possibles.

%%%%%%%%%%%%%%%%%
\section{Complexité ontologique}
%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%
%\subsection{Multidimensionalité et diversité d'approches}

Une première approche théorique permet une entrée sur ce que nous appelons \emph{complexité ontologique}, qui a été proposée par \cite{pumain2003approche} comme la largeur des points de vue disciplinaires nécessaires pour appréhender un même objet. La multidimensionalité des systèmes territoriaux reste un enjeu principal pour leur compréhension, comme l'illustre \cite{perez2016agent} dans le cas des systèmes multi-agents. Ainsi, l'interdisciplinarité serait intrinsèque aux ``sciences urbaines'', c'est-à-dire aux différentes disciplines prenant la ville comme objet d'étude, même si celle-ci s'avère en pratique laborieuse et soumise à la contingence de l'organisation sociale des sciences \citep{dupuy2015sciences}.

Nous reprenons l'exemple de \cite{raimbault2017invisible} comme preuve-de-concept de la diversité des approches possibles dans le cas des relations entre réseaux et territoires, et suggérons des pistes de réflexion quant au rôle de l'espace dans cette complexité, comme les processus évolutifs de diversification ou de spécialisation liés aux niches spatiales. Plus précisément, \cite{raimbault2017invisible} établit une cartographie scientifique des approches de modélisation des interactions entre réseaux de transport et territoires. Par une étude de réseaux de citations, émergent des approches complémentaires, incluant par exemple la géographie quantitative, la géographie politique, l'économie urbaine, la planification, la physique. Chaque approche éclaire sur une dimension particulière du même système. L'espace permettant la différentiation des sous-systèmes, il parait d'une part naturel que certains soient plus saillants que d'autres sur les différentes instances, et d'autre part que des interrogations complémentaires cohabitent (celles-ci étant en fait endogènes aux systèmes territoriaux, puisque l'innovation et donc la recherche et la science sont un moteur important des dynamiques des systèmes urbains \citep{pumain2010theorie}).

%%%%%%%%%%%%%%%%%
%\subsection{L'espace porteur de richesse ontologique ?}

À ce point, nous proposons une hypothèse, dont l'exploration empirique nécessiterait des analyses scientométriques poussées hors de portée de ce travail, selon laquelle une spatialisation plus élaborée serait lié à un éventail ontologique plus large. L'exemple de la \emph{New Economic Geography} et de la géographie économique illustre ce cas \citep{marchionni2004geographical}: la première approche, afin de déployer ses outils analytiques, impose un réductionnisme puissant sur l'espace (ligne ou cercle pour une grande partie des modèles) et sur les objets (agents représentatifs, homogénéité), tandis que la seconde favorisera des descriptions empiriques fidèles aux particularités géographiques. Il est difficile de dire si l'espace est plus riche parce que l'approche n'est pas réductionniste ou le contraire, qu'un espace riche augmente la portée ontologique. Prétendre un sens de causalité serait en fait contre-productif, et cette congruence confirme notre argument qu'une spatialisation élaborée des modèles des systèmes territoriaux va de pair avec une plus grande richesse ontologique.

Cette reflexion peut être poussée sur le plan méthodologique, au sein duquel on peut retrouver la tension entre précision du modèle et robustesse des résultats, notamment en comparant les modèles basés-agent et les modèles de systèmes dynamiques permettant un certain niveau de résolution analytique. L'archétype des approches en géographie économique, considérant un système unidimensionnel \citep{krugman1992dynamic}, perd l'essence de l'espace qui est hétérogénéité et multiplicité des alternatives, tandis que des modèles précisément spatialisés comme les modèles Luti, nécessiteront simulation et une validation restreinte des résultats \citep{bonnel2014survey}. Le degré de précision des modèles n'est cependant pas directement relié à leur caractère spatial, comme le montre la comparaison entre modèles agents et microsimulation démographique \citep{birkin2011spatial}. Cependant, les approches incluant une description élaborée de l'espace, avec la prise en compte des échelles par exemple, sont principalement associées à une richesse ontologique dans l'étude de ces systèmes, qu'il convient d'entretenir pour maintenir la diversité des connaissances sur le sujet \citep{sanders2018survival}.

%%%%%%%%%%%%%%%%
\section{Complexité et émergence}
%%%%%%%%%%%%%%%%

Notre deuxième entrée théorique s'intéresse à la complexité en tant qu'émergence faible de structures et autonomie des niveaux supérieurs~\citep{bedau2002downward}. L'émergence faible consiste en l'interprétation que les niveaux d'un système sont effectivement \emph{computés} \citep{morin1980methode} par celui-ci, et qu'il est nécessaire de simuler le système pour les reproduire. L'aspect diachronique des dynamiques permet que cette vue de l'émergence ne soit pas incompatible avec la causalité descendante. Nous rappelons le caractère intrinsèquement multi-échelle des systèmes territoriaux, qui se manifeste par exemple dans l'approche des villes comme systèmes au sein de systèmes de villes, comme l'introduit \cite{pumain1997pour} qui enrichit \cite{berry1964cities}. Par ailleurs, il existe une nécessité actuelle de production de modèles spatiaux intégrant effectivement cet aspect multi-échelle~\citep{rozenblat2018conclusion}, dans le but de modèles effectivement opérationnels.

La difficulté d'endogénéisation de niveau supérieurs autonomes peut par exemple être illustrée par~\cite{lenechet:halshs-01272236} qui propose un modèle de co-évolution entre transport et usage du sol à l'échelle métropolitaine intégrant une structure de gouvernance endogène pour le réseau de transport. Simulant les négociations entre acteurs locaux du transport, certains régimes conduisent à l'émergence d'un niveau intermédiaire de gouvernance issu de la collaboration entre acteurs voisins. Les trois niveaux décisionnels sont alors bien autonomes ontologiquement mais aussi en termes de dynamiques. L'émergence de la collaboration est finement liée à la structure spatiale, puisque les acteurs incluent les motifs d'accessibilité dans leur processus de prise de décision. Des modèles moins riches de la co-évolution entre transport et usage du sol, comme \cite{raimbault2018urban} à l'échelle mesoscopique prenant en compte forme urbaine et topologie du réseau routier, ou \cite{raimbault2018modeling} qui abstrait les réseaux sous formes de matrices de distance à l'échelle macroscopique et postule des dynamiques sur celles-ci, capturent bien une émergence fondamentalement liée à l'espace, mais restent restreints dans l'articulation des niveaux, tandis que le modèle de gouvernance est plus proche de l'émergence qualitative de niveaux supérieurs.

Des systèmes de complexité ontologique moindre présentent aussi un rôle crucial de l'émergence faible. L'état local d'un flux de trafic est en partie conséquence de l'état global du système, en particulier lorsque des motifs de congestion conséquents sont observables à l'échelle macroscopique. Dans ce cas, les motifs spatio-temporels sont encore cruciaux dans le processus d'émergence \citep{treiber2010three}. Dans ces différents exemples, l'espace joue encore un rôle crucial pour la présence de complexité, puisque l'auto-organisation implique agencement spatial des agents, et les niveaux d'émergence sont directement liés aux échelles spatiales quel que soit l'interprétation de celles-ci \citep{manson2008does}. Une approche de l'auto-organisation par la morphogenèse, que nous détaillerons au sein de modèles par la suite, insiste d'autant plus sur l'aspect spatial dans l'émergence, puisqu'elle fait le lien entre forme et fonction \citep{doursat2012morphogenetic}, la première étant nécessairement spatialisée.

%%%%%%%%%%%%%%%%%
\section{Complexité dynamique}
%%%%%%%%%%%%%%%%%

Dans cette section ainsi que la suivante, nous proposons d'utiliser des modèles de simulation de morphogenèse urbaine pour montrer de manière concrète d'autres complexités des systèmes territoriaux. Les modèles, détaillés par la suite, sont implémentés en Netlogo \citep{wilensky1999netlogo} et en scala, et explorés à l'aide du logiciel d'exploration de modèles OpenMOLE \citep{reuillon2013openmole}. L'ensemble du code et des résultats est disponible de manière ouverte sur le dépôt git du projet à \texttt{https://github.com/JusteRaimbault/SpatialComplexity}. Les résultats de simulation sont disponibles à \texttt{https://doi.org/10.7910/DVN/LENFVH}.

%\subsection{Systèmes dynamiques, chaos et fractales}

La compréhension des systèmes complexes comme systèmes dynamiques aux attracteurs plus ou moins chaotiques a été largement développée en géographie \citep{dauphine1995chaos}. Par exemple, \cite{e18060197} considère que l'information sémantique d'un environnement urbain est liée aux attracteurs de systèmes dynamiques régissant les dynamiques de ses agents. Ce type d'approche peut par ailleurs être rapprochée des approches fractales des systèmes urbains, initialement introduites par \cite{batty1994fractal}, et par exemple plus récemment appliquées à des problèmes réels de planification urbaine \citep{yamu2015spatial}. Ces questions sont liées plus généralement à des problématiques transversales de chaos spatio-temporel \citep{crutchfield1987phenomenology}. La compréhension du lien entre temps et espace, et des dynamiques non-uniformes, non-stationnaires, non-ergodiques correspondantes, reste en construction sur les plans théoriques, méthodologique et empirique, et promet de nombreuses applications pour l'étude des systèmes territoriaux. Par exemple,  \cite{chen2009urban} combine les correlations croisées et l'analyse de Fourier pour étudier un modèle de gravité urbaine. Une direction de recherche importante dans ce cadre est la compréhension de la non-stationnarité des propriétés des systèmes territoriaux, et \cite{raimbault2018urban} l'explore dans le cas de la morphologie urbaine et de la forme des réseaux.

%\subsection{Sensibilité aux conditions initiales et dépendance au chemin}

Nous utilisons ici un modèle de morphogenèse urbaine introduit par \cite{raimbault2018calibration} pour illustrer les propriétés de sensibilité aux conditions initiales et de dépendance au chemin des systèmes territoriaux~\citep{pumain2012urban}. En particulier, nous montrons la forte sensibilité des formes urbaines finales simulées aux perturbations spatiales, et plus généralement la dépendance au chemin des trajectoires pour les indicateurs morphologiques agrégés. Ce modèle relativement simple simule la croissance de population distribuée sur une grille, par ajout itératif de population, qui se localise suivant un attachement préférentiel à la population déjà présente puis se diffuse dans l'espace. Les paramètres cruciaux sont $N_G$ le taux de croissance exogène, $P_{max}$ la population totale finale, $\alpha$ l'exposant d'attachement préférentiel, $\beta$ le taux de diffusion et $n_d$ un nombre de diffusions par pas de temps.

L'experience menée par \cite{raimbault2018calibration} sur une version simplifiée à une dimension du modèle montre que les distributions semi-stationnaires de population peuvent être à distance maximale (au sens d'une norme L2 entre les distributions) à partir d'une même configuration initiale. En deux dimensions, le phénomène est identique comme l'expérience menée ici le montre. Le modèle est simulé à partir d'une configuration initiale particulièrement sensible, constituée de 4 centres initiaux de taille équivalente. Pour une grille de taille 100, 4 centres sont positionnés au milieu de chaque cadrant, avec un noyau exponentiel de population de la forme $P_0 \cdot \exp \left(-r/r_0\right)$ avec $P_0 = 100$ et $r_0 = 5$. Le modèle est alors simulé pour des valeurs données des paramètres d'agrégation $\alpha$, de diffusion $\beta, n_d$, de croissance et population $N_G, P_{max}$ (voir \cite{raimbault2018calibration}), et une réalisation de la distance entre configuration $d(t)$ est calculée sur deux réalisations indépendantes des populations $P^{(k)}_i(t)$, comme $d(t)=\norm{P^{(1)}_i(t) - P^{(2)}_i(t)}$. Un plan d'expérience direct est effectué, par échantillonnage LHS de 390 points de paramètres et 1000 répétitions.

@@

Nous montrons en Fig.~\ref{fig:lyapounov} une estimation basique des exposants de Lyapounov, effectuée par ajustement par régression linéaire par morceaux avec deux segments, de $\log d(t)$ en fonction de $t$, sur l'ensemble des répétitions pour un point de paramètres. L'ajustement est relativement bon puisque le premier quartile du $R^2$ ajusté est à 0.92 et sa moyenne à 0.93. L'exposant de Lyapounov sur le premier segment, qu'on note $\lambda_1$, varie entre 0.05 et 0.84 avec une moyenne de 0.31. Ainsi, la qualité de l'ajustement et les valeurs positives montrent la forte sensibilité aux conditions initiales des configurations générées. Les régimes temporels sont assez différents, puisque le temps de rupture $t_b$ estimé varie entre 2.5 et 37.0. Les valeurs de $\lambda_2$ sont plus faibles, avec une moyenne à 0.02. Il existe un premier régime durant lequel le principal de la structure urbaine se constitue puis un deuxième régime (qu'on pourrait qualifier de ``semi-stationnaire'') pendant lequel les évolutions sont plus mineures. Cela suggère par ailleurs un critère d'arrêt endogène au modèle. Les courbes de $\log d(t)$ montrées en Fig.~\ref{fig:lyapounov} sont pour une valeur maximale de l'ajustement et une valeur maximale de $\lambda_1$. Nous montrons aussi les variations des exposants en fonction des paramètres. $\lambda_1$ en fonction de l'agrégation $\alpha$ présente un comportement inattendu, relativement stable en dessous de 1.5 puis une croissance linéaire au dessus, sans modification autour de 1, ce qui suggère un seuil au dessus duquel ce processus influe drastiquement sur la forme urbaine. Pour les plus faibles valeurs de $\beta$, un maximum local à 1 suggère un rôle de l'aspect infra ou supra-linéaire du régime. Pour le comportement de $\lambda_2$, un effet significatif s'observe uniquement pour les plus forts taux de croissance, et correspond à un maximum autour de 1.5, c'est-à-dire que le passage au second régime pour $\lambda_1$ correspond dans ce cas à une meilleure stabilisation pour la seconde phase. Cette étude montre donc la sensibilité des configurations urbaines à des perturbations infinitésimales (ce qui est est équivalent à une perturbation de l'état initial).

%%%%%%%%%%%%%
\begin{figure}
	\includegraphics[width=0.49\linewidth]{figures/configdist_boxplot_id3642.png}
	\includegraphics[width=0.49\linewidth]{figures/configdist_boxplot_id3784.png}
	\includegraphics[width=0.49\linewidth]{figures/lambda1_alpha_colbeta.png}
	\includegraphics[width=0.49\linewidth]{figures/lambda2_alpha_colrelgrowthrate.png}
	\caption{\textbf{Estimation des exposants de Lyapounov locaux pour le modèle de Réaction-diffusion.} \textit{(Haut)} Profils temporel sous forme de statistiques descriptives pour $\log d(t)$ en fonction de $t$, pour deux configurations présentant un ajustement maximal et une valeur de $\lambda_1$ maximale ; \textit{(Bas gauche)} Valeur de $\lambda_1$ en fonction de $\alpha$, pour différentes valeurs de $\beta$ (couleur) ; \textit{(Bas droite)} Valeur de $\lambda_2$ en fonction de $\alpha$, pour différentes valeurs de $N_G / P_{max}$.}
	\label{fig:lyapounov}
\end{figure}
%%%%%%%%%%%%%

>>>>

\title{Space and complexities of territorial systems
\\\bigskip
\textit{Actes des Journ{\'e}es de Rochebrune 2019}
}
\author{\noun{Juste Raimbault}$^{1,2,3}$
\bigskip\\
$^1$ UPS CNRS 3611 ISC-PIF\\
$^2$ CASA, UCL\\
$^3$ UMR CNRS 8504 G{\'e}ographie-cit{\'e}s
}
\date{}

@@

\begin{abstract}
	The spatial character of territorial systems plays a crucial role in the emergence of their complexities. This contribution aims at illustrating to what extent different types of complexities can be exhibited in models of such systems. We develop from a theoretical viewpoint some arguments illustrating ontological complexity, in the sense of the diversity and multidimensionality of possible representations, and then complexity in the sense of emergence, i.e. the necessity of the existence of several autonomous levels. We then propose numerical experiments to explore properties of complexity (dynamical complexity and co-evolution) within two simple models of urban morphogenesis. We finally suggest other dimensions of complexity which could be typical of territorial systems.\medskip\\
	\noindent\textbf{Keywords: }\textit{Complexities; Territorial systems; Urban morphogenesis; Co-evolution}
\end{abstract}

@@

Territorial systems, which can be understood as self-organized socio-spatial structures  \citep{pumain1997pour}, are an archetypal of complex systems studied by numerous approaches which include more or less the spatial aspect of these systems. Our interpretation of territorial systems is more particularly situated within the heritage of the evolutionary urban theory~\citep{pumain2018evolutionary} which understands urban systems as multi-level systems, in which the co-evolution of multiple components and agents determine their dynamics~\citep{raimbault2018caracterisation}. The complexity of these systems is thus closely linked to their spatial nature, since their dynamics are carried within the spatial distribution of their entities and sub-systems, and interactions between agents causing the emerging behavior are embedded in space.

The concept of complexity corresponds to various approaches and definitions, which depend on the domains in which they were introduced. For example, \cite{chu2008criteria} reviews the conceptual and operational approaches linked to the domain of artificial life and shows that there is no unified concept. \cite{deffuant2015visions} also develops different visions of complexity, ranging from a complexity close to a computational complexity to an irreducible complexity typical of social systems. A previous work by \cite{raimbault2018relating} has proposed to suggest epistemological bridges between different approaches and definitions of complexity, and more particularly the links between complexity in the sense of emergence, computational complexity and informational complexity. As highlighted by \cite{batty2018defining}, dimensions of complexity are potentially infinite and an approach which would be universally valid for all problems and systems can not exist. It is therefore crucial to combine different approaches of complexity to understand its role within territorial systems.

This contribution aims at illustrating an approach of territorial systems through urban geography, and to what extent their complexity is closely linked to their spatial nature. We propose here to illustrate the links between complexity and space according to different views of it, both through theoretical considerations but also through the exploration of simulation models of territorial systems. The rest of this paper is structured as follows: first in two sections we develop a conceptual approach to ontological complexity and emergence in territorial systems. We then describe results from the simulation of an urban morphogenesis model which exhibits properties typical of dynamical complexity. An other experiment allows then to show the link between complexity in the sense of a co-evolution and the spatial structure of the system. We finally discuss the implications of these results and possible developments.

%%%%%%%%%%%%%%%%%
\section{Ontological complexity}
%%%%%%%%%%%%%%%%%

A first theoretical approach allows to tackle what we call \emph{ontological complexity}, which was proposed by \cite{pumain2003approche} as the broadness of disciplinary viewpoints necessary to apprehend a common object. The multidimensionality of territorial systems remains a main issue for their understanding, as show \cite{perez2016agent} in the case of multi-agent systems. Thus, interdisciplinarity would be intrinsic to ``urban sciences'', i.e. to the different disciplines considering the city as an object of study, although it is in practice difficult and subject to the contingency of the social organisation of sciences \citep{dupuy2015sciences}.

We take the example of \cite{raimbault2017invisible} as a proof-of-concept of the diversity of approaches possible in the case of interactions between networks and territories, and suggest hypotheses regarding the role of space in this complexity, such as the evolutive processes of diversification and specialization linked to spatial niches. More precisely, \cite{raimbault2017invisible} establishes a scientific map of modeling approaches to interaction between transportation networks and territories. Through a citation network analysis, complementary approaches are exhibited, including for example quantitative geography, political geography, urban economics, planning, physics. Each approach provides a particular dimension of the same system. Space leading to the differentiation of sub-systems, it seems on the one hand natural that some have a higher importance than others in the different instances, and on the other hand that complementary questionings coexist (these being in fact endogenous to territorial systems, since innovation and thus research and science are an important driver of urban systems dynamics \citep{pumain2010theorie}).

At this point we propose an hypothesis, which empirical exploration would require deeper scientometric analysis out of the scope of this work, according to which a more elaborated spatialization would be linked to a broader ontological spectrum. The example of \emph{New Economic Geography} and geographical economics illustrates that aspect \citep{marchionni2004geographical}: the first approach, in order to deploy its analytical tools, imposes a strong reductionism on space (line or circle for a large part of models) and on objects (representative agents, homogeneity), whereas the second will favor empirical descriptions more faithful to geographical particularities. It is difficult to know if space is richer because the approach is not reductionist or the contrary, that a rich space increases ontological range. Pretend a sense of causality would in fact be counter-productive, and this congruence confirms our argument that an elaborated spatialization of models of territorial systems is correlated to a higher ontological richness.

This reflexion can be shifted to the methodological domain, within which we can highlight the tension between model precision and robustness of results, in particular by comparing agent-based models and dynamical systems models allowing a certain level of analytical tractability. The archetype of approaches in economic geography, considering an unidimensional system \citep{krugman1992dynamic}, looses the essence of space which is heterogeneity and multiplicity of alternatives, whereas precisely spatialized models such as Luti models, will necessitate simulations and a restricted validation of results \citep{bonnel2014survey}. The degree of precision of models is however not directly linked to their spatial character, as shows the comparison between agent-based models and demographic microsimulations \citep{birkin2011spatial}. However, approaches including an elaborated description of space, with the inclusion of scales for example, are mainly associated to an ontological richness in the study of these systems, which need to be maintained to ensure the diversity of knowledges on the subject \citep{sanders2018survival}.

%%%%%%%%%%%%%%%%
\section{Complexity and emergence}
%%%%%%%%%%%%%%%%

Our second theoretical entry focuses on complexity as the weak emergence of structures and autonomy of upper levels~\citep{bedau2002downward}. Weak emergence consists in the interpretation that the levels of a systems are effectively \emph{computed} \citep{morin1980methode} by itself, and that it is necessary to simulate the system to reproduce them. The diachronic aspect of dynamics allows this view of emergence to not being incompatible with downward causation. We recall the intrinsic multi-scalar aspect of territorial systems, which is highlighted for example in the approach of cities as systems within systems of cities, as introduced by \cite{pumain1997pour} which extends \cite{berry1964cities}. Furthermore, there is currently a necessity to produce spatial models integrating effectively this multi-scalar aspect~\citep{rozenblat2018conclusion}, with the aim of effectively operational models.

The difficulty to endogenize autonomous upper levels can for example be illustrated by~\cite{lenechet:halshs-01272236} which introduces a co-evolution model between transportation and land-use at the metropolitan scale integrating an endogenous governance structure for the transportation network. Simulating tractations between local actors of transportation, some regimes lead to the emergence of an intermediary governance level produced by the collaboration between neighbor actors. The three decision levels are then indeed ontologically autonomous but also in terms of their dynamics. The emergence of collaboration is finely linked to the spatial structure, since actors include accessibility patterns in their decision-making process. Less rich models of co-evolution between transport and land-use, as \cite{raimbault2018urban} at the mesoscopic scale taking into account urban form and topology of the road network, or \cite{raimbault2018modeling} which abstract networks as distance matrices at the macroscopic scale and postulates dynamics on these, indeed capture an emergence fundamentally linked to space, but remain restricted in the articulation of levels, whereas the governance model is closer to the qualitative emergence of upper levels.

Systems with a lower ontological complexity also exhibit a crucial role of weak emergence. The local state of a traffic flow is partly a consequence of the global state of the system, in particular when important congestion patterns can be observed at the macroscopic scale. In that case, spatio-temporal patterns are again central to the emergence process \citep{treiber2010three}. In these different examples, space again play a crucial role for the presence of complexity, since auto-organization implies a spatial distribution of agents, and the levels of emergence are directly linked to spatial scales whatever the interpretation of these \citep{manson2008does}. An approach of auto-organization through morphogenesis, that will be detailed within models in the following, furthermore insists on the spatial aspect in emergence, since it links form and function \citep{doursat2012morphogenetic}, the first being necessarily spatialized.

%%%%%%%%%%%%%%%%%
\section{Dynamical complexity}
%%%%%%%%%%%%%%%%%

In this section and the following, we propose to use simulation models of urban morphogenesis to show in a concrete manner other complexities of territorial systems. The models, detailed in the following, are implemented in Netlogo \citep{wilensky1999netlogo} and in scala, and explored using the model exploration software OpenMOLE \citep{reuillon2013openmole}. All code and results is openly available on the git repository of the project at \texttt{https://github.com/JusteRaimbault/SpatialComplexity}. Simulation results are available at \texttt{https://doi.org/10.7910/DVN/LENFVH}.

The understanding of complex systems as dynamical systems with more or less chaotic attractors has been largely developed in geography \citep{dauphine1995chaos}. For example, \cite{e18060197} considers that the semantic information of an urban environment is linked to the attractors of dynamical systems ruling the dynamics of its agents. This type of approach can moreover be linked to fractal approaches of urban systems, initially introduced by \cite{batty1994fractal}, and for example more recently applied to real urban planning problems \citep{yamu2015spatial}. These questions are more generally linked to transversal questions on spatio-temporal chaos \citep{crutchfield1987phenomenology}. The understanding of the link between time and space, and the corresponding non-uniform, non-stationary, non-ergodic dynamics, remains in construction on the theoretical, methodological and empirical dimensions, and promises numerous applications for the study of territorial systems. For example, \cite{chen2009urban} combines cross-correlations and Fourier analysis to study an urban gravity model. An important research direction in that frame is the understanding of the non-stationarity of territorial systems properties, and \cite{raimbault2018urban} explores it in the case of urban morphology and network topology.

% TODO : ajouter edge of chaos ?

We use here an urban morphogenesis model introduced by \cite{raimbault2018calibration} to illustrate the properties of sensitivity to initial conditions and path-dependancy of territorial systems~\citep{pumain2012urban}. In particular, we show the strong sensitivity of final simulated urban forms to spatial perturbations, and more generally the path-dependancy of trajectories for aggregated morphological indicators. This relatively simple model simulates population growth on a grid, through the iterative addition of population, which localizes following a preferential attachment to the already distributed population and diffuses in space. Important parameters are $N_G$ the exogenous growth rate, $P_{max}$ the total final population, $\alpha$ the preferential attachment exponent, $\beta$ the diffusion rate and $n_d$ a number of diffusions for each time step.

The experiment done by \cite{raimbault2018calibration} on a simplified one-dimensional version of the model shows that semi-stationary distributions of populations can be at a maximal distance (in a sense of a L2 norm between distributions) starting from a same initial configuration. In two dimensions, the phenomenon is identical as shown by the experiment done here. The model is simulated starting from an initial configuration which is particularly sensitive, constituted by four initial centers of similar size. For a grid of size 100, four centers are positioned in the middle of each cadrant, with an exponential kernel for population of the form $P_0 \cdot \exp \left(-r/r_0\right)$ with $P_0 = 100$ and $r_0 = 5$. The model is then simulated for given values of parameters for aggregation $\alpha$, diffusion $\beta, n_d$, growth and population $N_G, P_{max}$ (voir \cite{raimbault2018calibration}), and a realization of the distance between configurations $d(t)$ is computed on two independent realizations of populations $P^{(k)}_i(t)$, as $d(t)=\norm{P^{(1)}_i(t) - P^{(2)}_i(t)}$. A direct experience plan is achieved, with a LHS sampling of 390 parameter points and 1000 repetitions.

@@

We show in Fig.~\ref{fig:lyapounov} a basic estimation of Lyapounov exponents, achieved with piecewise linear regression estimation on two segments, of $\log d(t)$ as a function of $t$, on all repetitions for one parameter point. The adjustment is relatively good since the first quartile of the adjusted $R^2$ is at 0.92 and its average at 0.93. The Lyapounov exponent on the first segment, that we write $\lambda_1$, varies between 0.05 and 0.84 with an average of 0.31. Thus, the quality of the adjustment and the positive values show the strong sensitivity to initial conditions of generated configurations. Temporal regimes are quite different, since the estimated breakpoint time $t_b$ varies between 2.5 and 37.0. Values of $\lambda_2$ are lower, with an average at 0.02. There exists a first regime during which the most of the urban structure is constituted followed by a second regime (that we can describe as ``semi-stationary'') during which evolutions are smaller. This moreover suggests an endogenous stopping criteria for the model. The curves of $\log d(t)$ shown in Fig.~\ref{fig:lyapounov} are for a maximal value of the adjustment and a maximal value of $\lambda_1$. We also show the variations of exponents as a function of parameters. $\lambda_1$ as a function of the aggregation $\alpha$ exhibits an unexpected behavior, relatively stable below 1.5 and then a linear increase above, without modification around 1, what suggests a threshold over which this process dramatically influences urban form. For the lower values of $\beta$, a local maximum at 1 suggests a role of the infra or supra-linear aspect of the regime. For the behavior of $\lambda_2$, a significant effect is observed only for the larger growth rates, and corresponds to a maximum around 1.5, i.e. that the switch to the second regime for $\lambda_1$ corresponds in this case to a better stabilization for the second phase. This study shows thus the sensitivity of urban configurations to infinitesimal perturbations (what is equivalent to a perturbation of the initial state).

%%%%%%%%%%%%%
\begin{figure}
	\includegraphics[width=0.49\linewidth]{figures/configdist_boxplot_id3642.png}
	\includegraphics[width=0.49\linewidth]{figures/configdist_boxplot_id3784.png}
	\includegraphics[width=0.49\linewidth]{figures/lambda1_alpha_colbeta.png}
	\includegraphics[width=0.49\linewidth]{figures/lambda2_alpha_colrelgrowthrate.png}
	\caption{\textbf{Estimation of local Lyapounov exponents for the reaction-diffusion model.} \textit{(Top)} Temporal profiles as descriptive statistics for $\log d(t)$ as a function of $t$, for two configurations showing a maximal fit and a maximal $\lambda_1$ value ; \textit{(Bottom left)} Value of $\lambda_1$ as a function of $\alpha$, for different values of $\beta$ (color) ; \textit{(Bottom right)} Value of $\lambda_2$ as a function of $\alpha$, for different values of $N_G / P_{max}$.}
	\label{fig:lyapounov}
\end{figure}
%%%%%%%%%%%%%

Furthermore, we illustrate the temporal trajectories of aggregated morphological indicators. A similar experience plan with a single model simulation at each repetition and an empty initial state, but following in time the values of urban form indicators (Moran, entropy, average distance, hierarchy), shows that their trajectories are path-dependent. Indeed, the example of Fig.~\ref{fig:morphotraj} shows a significant divergence of trajectories starting from the same initial point, but more importantly some trajectories crossing (indicators may therefore not be formulated as a deterministic dynamical system in which trajectories may not cross because of the Cauchy-Lipschitz theorem), exhibiting numerous points where urban form is very close (at least for the two indicators shown) for two trajectories, but the past and the future are different. Therefore, the future state of an urban system entirely depends on the past trajectory, and two systems which are very similar at a given time may exhibit fundamentally different trajectories. It is naturally an illustration with a toy model, on the particular aspect of urban form, but the path-dependence effects are a priori even more important when taking into account the multiple levels, political aspects, and infrastructures \citep{pumain2012urban}.

%%%%%%%%%%%%%
\begin{figure}
	\centering
	\includegraphics[width=0.7\linewidth]{figures/trajs_moran-dist_seed8578.png}
	\caption{\textbf{Temporal trajectories of the morphological indicators.} We show for a parameter point a sample of 50 trajectories in the morphological space (moran,distance), the color giving the level of hierarchy. In this configuration, the final variance is very large and trajectories sometimes cross others, confirming the path-dependence for morphological indicators.}
	\label{fig:morphotraj}
\end{figure}
%%%%%%%%%%%%%
